Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-12 talk/dls2012/paper.tex
correct the wrong depiction of luajit

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we present a scheme for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves. The scheme can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

A dynamic language typically needs to do quite a lot of type
checking, wrapping/unwrapping of boxed values, and virtual method dispatching. 
For tight computationally intensive loops a
significant amount of the execution time might be spent on such tasks
instead of the actual computations. Moreover, the type checking,
unwrapping and method lookups are often loop invariant and performance could be increased
by moving those operations out of the loop. We propose a simple scheme
to make a tracing JIT loop-aware by allowing it's existing optimizations to
perform loop invariant code motion.

@@

In this paper we want to address this problem by proposing a scheme that
makes it possible to turn optimizations using one forward pass into
optimizations that can do loop invariant code motion and similar loop-aware
improvements. Using this scheme one does not need to change the underlying
optimization much to get these advantages.

@@

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, we want to get this effect using our existing optimization passes
without changing them too much. Optimizations with one forward pass
cannot directly get this effect: They just look at the trace without taking
into account that the trace executes many times in a row. Therefore to achieve
loop-invariant code motion, we peel one iteration off the loop before running
the optimizations. This peeling gives the following trace:

@@

This is the key insight of the proposed implementation scheme: If an
optimization is given two iterations together at the same time, the
optimization has enough context to remove operations from the peeled loop,
because it detects
that the operation was performed in the preamble already. Thus at runtime these
moved operations are only executed once when entering the loop and the results
are reused in further iterations.

@@

Before a trace is passed to the backend compiling it into machine code
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop making them executed only once
and not every iteration. We propose to achieve this by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimizations of interest only a few
additional details have to be considered when they are combined with loop peeling. These are
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

$L_1$($p_{0}$, $p_{5}$):
# inside f: y = y.add(step)
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = $i_{6}+i_{7}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($L_1$, $p_{0}$, $p_{9}$)
\end{lstlisting}
\caption{A peeled trace of the Example Interpreter}
\label{fig:peeled-trace}
\end{figure}

@@

\revd{
The benchmark results appear quite impressive -- especially the comparison with
GCC -- but without additional information, I have no idea what is being
compared.  Are these results from the same sizes of integers and/or floating
point results?
}

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks. For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C.

@@

We run GCC both with -O2 optimization and -O3 -march=native, disabling the
automatic loop vectorization. In all cases, SSE2 instructions were used for
floating point operations, except Psyco which uses x87 FPU instructions.
We also run PyPy with loop peeling optimization and without (but otherwise
identical).

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT assembler backend as well as missing
optimizations, like instruction scheduling.

@@

Loop invariant code motion optimizations are completely
standard~\cite{muchnick_advanced_1997}. Therefore, the effects that our
optimization achieves are not in any way new. However, we think that achieving
them in the way described in this paper is simpler than writing explicit algorithms.

\revc{
The discussion of LuaJIT is unsatisfying.  It's not clear to me from that one
quote that Mike is doing the same thing.  It might be worth including LuaJIT in
the benchmarks, and/or examining the actual implementation of LuaJIT.
}
\cfbolz{maybe we can look in the new LuaJIT wiki.
how annoying would it be to rerun the benchmarks, if I can find somebody to write them?}
\hakan{there is iwtc11/benchmarks/runall.sh which is supposed to run them all}

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developed the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is
generated.''~\cite{pall_luajit_2009}

@@

In this paper we have studied loop invariant code motion during trace
compilation. We claim that loop peeling is a very convenient solution
here since it fits well with other trace optimizations and does not require
large changes to them. This approach improves the effect of standard
optimizations such as redundant guard removal, common subexpression elimination
and allocation removal. The most prominent effect is that they all become loop
invariant code motion optimizations.

@@

\acks
We would like to thank Samuele Pedroni, Sven Hager and the anonymous reviewers
for helpful comments on drafts of this paper.

>>>>

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we explain a scheme invented within the context of the LuaJIT project
for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves.
We have implemented the scheme in PyPy's tracing JIT compiler,
where it can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

A dynamic language typically needs to do quite a lot of type
checking, wrapping/unwrapping of boxed values, and virtual method dispatching. 
For tight computationally intensive loops a
significant amount of the execution time might be spent on such tasks
instead of the actual computations. Moreover, the type checking,
unwrapping and method lookups are often loop invariant and performance could be increased
by moving those operations out of the loop. We explain a simple scheme
to make a tracing JIT loop-aware by allowing it's existing optimizations to
perform loop invariant code motion.

@@

Mike Pall pioneered a solution to address this problem in the context of a
dynamic language using a tracing JIT compiler. He published his algorithm and
its rationale in 2009~\cite{pall_luajit_2009} and implemented it in LuaJIT
2.0\footnote{\url{http://luajit.org/}}, an open source JIT compiler for the Lua
language. His approach allows to reuse all forward pass
optimizations to achieve loop invariant code motion and other loop-related
optimizations, which greatly simplifies the implementation. Using this scheme
one does not need to change the underlying optimization much to get these
advantages. We have implemented the same approach in PyPy's tracing JIT
compiler the results of which we present here.

@@

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, it is desirable to get this effect using our existing optimization passes
without changing them too much. Optimizations with one forward pass
cannot directly achieve this effect: They just look at the trace without taking
into account that the trace executes many times in a row. Therefore to achieve
loop-invariant code motion, we peel one iteration off the loop before running
the optimizations. This peeling gives the following trace:

@@

This is the key insight of the implementation scheme: If an
optimization is given two iterations together at the same time, the
optimization has enough context to remove operations from the peeled loop,
because it detects
that the operation was performed in the preamble already. Thus at runtime these
moved operations are only executed once when entering the loop and the results
are reused in further iterations.

@@

Before a trace is passed to the backend compiling it into machine code
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop making them executed only once
and not every iteration. This can be achieved by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimizations of interest only a few
additional details have to be considered when they are combined with loop peeling. These are
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

$L_1$($p_{0}$, $p_{5}$):
# inside f: y = y.add(step)
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = $i_{6}+i_{7}$
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($L_1$, $p_{0}$, $p_{9}$)
\end{lstlisting}
\caption{A peeled trace of the example interpreter}
\label{fig:peeled-trace}
\end{figure}

@@

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks. For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C. All the benchmarks except sqrt operate on C double-precision floating
point numbers, both in the Python and the C code.

@@

We run GCC with -O3 -march=native, disabling the
automatic loop vectorization. In all cases, SSE2 instructions were used for
floating point operations, except Psyco which uses x87 FPU instructions.
We also run PyPy with loop peeling optimization and without (but otherwise
identical).

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT machine code backend as well as missing
optimizations, like instruction scheduling.

@@

Loop invariant code motion optimizations are a well-known approach to optimize
loops~\cite{muchnick_advanced_1997}. Therefore, the effects that the
optimizations described here achieve are not in any way new. However, we think
that achieving them in the way described in this paper is simpler than writing
explicit algorithms.
\cfbolz{more explicit listing of prior work goes here}

As described in the introduction,
Mike Pall pioneered the approach described in this paper.
He showed that, unlike traditional loop-invariant code motion
(LICM), this approach is effective, even in the presence of many
guards and global control dependencies, which are caused by the
semantics of dynamic languages.

He writes on the Lua-users mailing list:
``The LOOP pass does synthetic unrolling of the recorded IR, combining
copy-substitution with redundancy elimination to achieve code hoisting. The
unrolled and copy-substituted instructions are simply fed back into the
compiler pipeline, which allows reuse of all optimizations for redundancy
elimination. Loop recurrences are detected on-the-fly and a minimized set of
PHIs is generated.''~\cite{pall_luajit_2009}

@@

In this paper we have studied loop invariant code motion during trace
compilation. We claim that the loop peeling approach of LuaJIT is a very convenient solution
since it fits well with other trace optimizations and does not require
large changes to them. The approach improves the effect of standard
optimizations such as redundant guard removal, common subexpression elimination
and allocation removal. The most prominent effect is that they all become loop
invariant code motion optimizations.

@@

\acks
We would like to thank Samuele Pedroni, Sven Hager and the anonymous reviewers
for helpful comments on drafts of this paper. We owe deep gratitude to Mike Pall
for making his impressive work on LuaJIT available and for detailed help on a
draft of the paper.
